---
title: "HW_morbidity_measures"
author: "Elena Marochkina"
date: "2024-09-15"
output:
  word_document: 
    toc: yes
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(glue)
```
\newpage
# Read Data

```{r}
data <- read_csv("data/carrental.csv")
glimpse(data)

```

# Calculate prevalence for experience
**Prevalence** is the number of people in the sample with the characteristic of interest (experience), divided by the total number of people in the sample.
```{r}
calculate_prevalence <- function(data, experience_flag) {
  prevalence <- data %>%
  summarize(
    prevalence_all = n(),
    prevalence = sum(experience == experience_flag)
  ) %>%
  mutate(prevalence = prevalence / prevalence_all) %>%
  pull(prevalence)
}

prevalence_experienced <- calculate_prevalence(data, 1)
prevalence_not_experienced <- calculate_prevalence(data, 0)

print(paste("Prevalence for experienced drivers is equal to", prevalence_experienced))
print(paste("Prevalence for non-experienced drivers is equal to", prevalence_not_experienced))


```
# Calculate incidence proportion(risk) of accident

**Risk** is the number of events (good or bad) in a treated (exposed) or control (nonexposed) group, divided by the number of people in that group.

```{r}
risk_calculation <- function(data, general_flag, experience_flag) {
  if (general_flag == TRUE) {
    # Calculate risk for the entire dataset
    risk_summary <- data %>%
      summarize(
        risk_all = n(),
        accident_count = sum(accident == 1)
      )
  } else {
    # Calculate risk for filtered data based on experience_flag
    risk_summary <- data %>%
      filter(experience == experience_flag) %>%
      summarize(
        risk_all = n(),
        accident_count = sum(accident == 1)
      )
  }

  risk_summary <- risk_summary %>%
    mutate(risk = (accident_count / risk_all) * 100)

  risk_value <- round(risk_summary$risk, 2)

  return(risk_value)
}

risk_general <- risk_calculation(data, TRUE, 0)
risk_experienced <- risk_calculation(data, FALSE, 1)
risk_not_experienced <- risk_calculation(data, FALSE, 0)

print(paste("Incidence proportion(Risk) for all drivers is equal to", risk_general, paste("%")))
print(paste("Incidence proportion(Risk) for experienced drivers is equal to", risk_experienced, paste("%")))
print(paste("Incidence proportion(Risk) for non-experienced drivers is equal to", risk_not_experienced, paste("%")))
```
# Calculate incidence rate of accident

**Incidence rate** is the ratio of the number of cases to the total time the population is at risk of disease.

```{r}

incidence_rate_calculation <- function(data, experience_flag = NA) {
  # Filter the data if experience_flag is provided
  if (!is.na(experience_flag)) {
    data <- data %>%
      filter(experience == experience_flag)
  }
  
  # Calculate observation time and accident count
  incidence_summary <- data %>%
    mutate(observation_time = stop - start) %>%
    summarize(
      total_observation_time = sum(observation_time),
      accident_count = sum(accident == 1)
    )
  
  # Calculate the total observation time in years
  total_years <- incidence_summary$total_observation_time / 365
  print(total_years)  # Optional: For debugging
  
  # Calculate incidence rate
  incidence_rate <- round(incidence_summary$accident_count / incidence_summary$total_observation_time, 4)
  incidence_rate_str <- paste(incidence_rate, "accidents per total observation time", round(total_years, 2), " years")
  
  # Calculate incidence rate per year
  incidence_rate_per_year <- round(incidence_summary$accident_count / total_years, 2)
  incidence_rate_per_year_str <- paste(incidence_rate_per_year, "accidents per year")
  
  return(list(incidence_rate = incidence_rate_str, incidence_rate_per_year = incidence_rate_per_year_str))
}

# Example usage:
incidence_rate_general <- incidence_rate_calculation(data)
incidence_rate_experienced <- incidence_rate_calculation(data, 1)
incidence_rate_not_experienced <- incidence_rate_calculation(data, 0)

print(paste("Incidence rate for all drivers is equal to", incidence_rate_general$incidence_rate, "or", incidence_rate_general$incidence_rate_per_year))
print(paste("Incidence rate for experienced drivers is equal to", incidence_rate_experienced$incidence_rate, "or", incidence_rate_experienced$incidence_rate_per_year))
print(paste("Incidence rate for non-experienced drivers is equal to", incidence_rate_not_experienced$incidence_rate, "or", incidence_rate_not_experienced$incidence_rate_per_year))

```
\newpage
# Summarise results and add logical explanation

## Results

1. Prevalence for experience is equal to **0.32**

2. Incidence proportion(Risk) for all drivers is equal to **22 %**

3. Incidence proportion(Risk) for experienced drivers is equal to **9.38 %**

4. Incidence proportion(Risk) for non-experienced drivers is equal to **27.94 %** per unit time (driver-years)

5. Incidence rate for all drivers is equal to **0.003** accidents per total observation time 19.79 years or **1.11** accidents per year

6. Incidence rate for experienced drivers is equal to **0.0032** accidents per total observation time 2.59 years or **1.16** accidents per year

7. Incidence rate for non-experienced drivers is equal to **0.003** accidents per total observation time 17.2 years or **1.1** accidents per year

## Overall Interpretation
1. **Prevalence** tells us that only 32 % of the drivers are experienced.

2. **Risk (Incidence Proportion)** shows that non-experienced drivers are significantly more likely to be involved in an accident than experienced drivers (27.94 % vs. 9.4 %).

3. **Incidence Rate** adjusts for the time of exposure, showing that experienced drivers have a slightly higher accident rate per unit of time than non-experienced drivers. This could suggest that the longer driving time for experienced drivers leads to more accidents, even though they are generally less risky when viewed as a percentage of the group.