---
title: "GitHub Task"
subtitle: "Biostatistics course: 1st task"
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output:
  word_document: 
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
library(readxl)
library(dplyr)
library(psych)
library(tibble)
library(tidyr)
library(stringi)
library(stringr)
library(flextable)
```

\newpage
# How to read and write data?

1. Read data from tsv file

```{r}
data_tsv <- read_delim("data/raw/data_tsv.tsv", delim = "\t", n_max = Inf)
```
2. Read data from tsv file and write data to csv file
```{r}
data_tsv2 <- read_tsv("data/raw/data_tsv.tsv", skip = 0, n_max = Inf)
write_excel_csv(data_tsv2, "data/raw/data_csv.csv")
```

3. Read data from csv file with , separator and write data to csv file with ; separator
```{r}
data_csv1 <- read_csv("data/raw/data_csv.csv")
write_csv2(data_csv1, "data/raw/data_csv2.csv")
```

4. Check if data doesn't change after format change
```{r}
isMatch <- all.equal(data_csv1, data_tsv2)
print(isMatch)
```

5. Read data from csv file with ; separator
```{r}
data_csv2 <- read_csv2("data/raw/data_csv2.csv")
```

6. Read data from excel file
```{r}
data_excel <- read_excel("data/raw/data_excel.xlsx", sheet = "data_csv2")
```

7. Write data in rds file
```{r}
write_rds(data_tsv2, "data/raw/data_rds.rds")
```

8. Read data from rds file
```{r}
data_rds <- read_rds("data/raw/data_rds.rds")
```

\newpage
# How to calculate and summarize basic statistics?

1. Mean calculation
```{r}
mean1 <- mean(c(1, -1, 5, -12, -12, 3, 8, -10, 0), trim = 0, na.rm = FALSE)
mean2 <- mean(c(-13, 19, -24, NA, 30, 64, -53, NA, 50, 31, -58, -34, -3, -34, 77), trim = 0, na.rm = TRUE)
```

2. Median calculation
```{r}
median1 <- median(c(1, 9, NA, 88, 2, NA, 42, NA, 4, 68, NA), na.rm = TRUE)
median2 <- median(c(-91, -33, 13, 34, 34, 75, -80, -35, -90, -72, 70, 67, -100, -94, -18), na.rm = FALSE)
```

3. Min max calculation
```{r}
arr1 <- c(48.11, 45.3, 58.42, 51.64, 62.07, 57.26, 49.69, 93.29, 81.18, 44.78, 55.1, 76.74, 58.08)
min <- min(arr1, na.rm = FALSE)
max <- max(arr1, na.rm = FALSE)
```

4. Quantile calculation
```{r}
quant1 <- quantile(c(80.94, 44.46, 46.33, 65.1, 66.42, 104.43, 53.15, 48.41, 12.88, 51.1, 43.03, 40.3, 33.71, 55.1, 22.17), probs = seq(0, 1, 0.25), na.rm = FALSE, names = TRUE, type =7)
quant2 <- quantile(c(26.17, 97.73, 24.81, 53.62, 87.72, 45.19, 45.7, 69.63, 36.76, 7.17), probs = seq(0, 1, 0.5))
```

5. Dispersion and SD calculation
```{r}
arr2 <- c(76.22, 65, 19.69, 29.84, 37.18, 70.93, 64.78, 61.66, 49.03, 51.56)
var <- var(arr2, na.rm = FALSE)
sd <- sd(arr2, na.rm = FALSE)
```

6. IQR calculation
```{r}
arr3 <- c(63.92, 35.85, 26.9, 48.92, 43.1, 66.94, 47.06, 56.54, 29.1, 58.88)
iqr <- IQR(arr3, na.rm = FALSE, type = 7)
```

7. Array length calculation
```{r}
len <- length(arr1)
len1 <- sum(!is.na(arr3))
```

8. Standard error of the mean calculation
```{r}
SEM <- function(x){
  result = sd(x) / sqrt(sum(!is.na(x)))
  return(result)
}
x <- c(76.22, 65, 19.69, 29.84, 37.18, 70.93, 64.78, 61.66, 49.03, 51.56)
SEM(x)
```

9. Convert categorical variables to factors
```{r}
data_tsv$Группа <- as.factor(data_tsv$Группа)
data_tsv$Пол <- as.factor(data_tsv$Пол)
```

10. Print summary of base statistics
```{r}
sum <- summary(data_tsv)
print(sum)
```

11. Print summary of base statistics for numeric data (second method). Psych packet
```{r}
numeric_data <- read_rds("data/raw/numeric_data.rds")
describe(numeric_data, na.rm = TRUE, skew = FALSE, ranges = TRUE)
```

12. Print summary of base statistics for categorical data (second method)
```{r}
categorical_data <- read_rds("data/raw/factor_data.rds")
table(categorical_data$Группа, categorical_data$`Группа крови`)
```

13. Print relative frequencies for categorical data
```{r}
prop.table(table(categorical_data$Группа, categorical_data$`Группа крови`))
```

\newpage
# How to work with Data Frames?
## Tibble package 
Get data: to read strings as factors must to use stringsAsFactors = TRUE 
```{r}
data <- read.csv("data/raw/data_csv.csv", stringsAsFactors = TRUE)
```

### Learning pipe
```{r}
data %>%
  select(., `Группа`, `Рост`, contains("E1")) %>%
  filter(., `Группа` == "Группа 1") %>%
  slice(., 1:10) %>%
  mutate(., `Рост` = `Рост` / 10)
```

### Variables access using tibble
```{r}
data$Группа
data$`Группа крови`
```

### Create variables using tibble
```{r}
tibble(var_first = 1:10, var_second = ifelse(var_first < 5, var_first + 100, var_first))
tibble(`var 1` = 1:10, `var 2` = `var 1` * 100)
```

### Data view in table format
```{r}
data %>% View()
```

### Add column or row
1. Add column
```{r}
data <- data %>%
  add_column(newHeight = data$`Рост` / 10, .after = Inf)
```

2. Add ow
```{r}
data <- data %>%
  add_row(`Группа` = "New_group", `Возраст` = 50, .before = 1)
```

### Add numeration
```{r}
data <- read.csv("data/raw/data_csv.csv", stringsAsFactors = TRUE)
data %>%
  mutate(ID = row_number())
```

## Binding
### Binding columns or rows
1. Binding columns
```{r}
data_1 <- tibble(ID_num = 1:10, Group = rep(c("placebo", "active"), 5))
data_2 <- tibble(random = rbeta(10, 3, 5), norm = rnorm(10))
data_3 <- tibble(Age = 100:91, Drugs = rep(c("omeprazole", "aspirin"), 5))
data_1 %>% bind_cols(data_2) %>% bind_cols(data_3)
```

2. Binding rows
```{r}
data_1 <- tibble(ID_num = 1, group = "active", test_res = "positive")
data_2 <- tibble(ID_num = 2, group = "placebo", test_res = "positive")
data_3 <- tibble(ID_num = 3, group = "placebo", test_res = "negative")
data_4 <- tibble(ID_num = 4, group = "active", test_res = "positive")

data_1 %>% bind_rows(data_2) %>% bind_rows(data_3) %>% bind_rows(data_4)
```

### Join table
```{r}
data_1 <- tibble(var_1 = 1:8) %>% mutate(id = row_number())
data_2 <- tibble(var_2 = rnorm(10)) %>% mutate(`Subject ID` = row_number())

data_1 %>%
  left_join(data_2, by = c("id" = "Subject ID"))
```
```{r}
data_1 %>%
  right_join(data_2, by = c("id" = "Subject ID"))
```
```{r}
data_1 %>%
  inner_join(data_2, by = c("id" = "Subject ID"))
```
```{r}
data_1 %>%
  full_join(data_2, by = c("id" = "Subject ID"))
```

## Group data
### Group by column or raw
1. Group by column
```{r}
data %>% group_by("Группа")
```

2. Group by row
```{r}
data %>% 
  rowwise() %>%
  mutate(`Average_hemoglobin` = mean(c(`Гемоглобин_E1`, `Гемоглобин_E2`))) %>%
  ungroup() %>%
  select(`Гемоглобин_E1`, `Гемоглобин_E2`, `Average_hemoglobin`)
```

### Divide data frame by values
```{r}
data %>% split("Группа")
```
## Selection
### Select columns
```{r}
data %>%
  select(`Группа.крови`, `Рост`)
```

```{r
}data %>% 
  select(!`Рост` & !`Пол`)
```

```{r}
data %>%
  select(where(is.numeric))
```

```{r}
data %>%
  select(where(is.factor))
```

```{r}
data %>% 
  select(`Группа`, (function(x) is.factor(x)) | where(function(x) is.numeric(x)))
```

#### Select data by visit 
```{r}
data %>%
  select(contains("_E1"))
```

```{r}
data %>%
  select(matches("_E\\d{1}"))
```

```{r}
data %>%
  select(starts_with("Б"))
```

#### Select strictly
```{r}
var <- c("Базофилы_E1", "Базофилы_E2", "Гемоглобин_E1")

data %>% 
  select(all_of(var))
```

#### Select gentle
```{r}
data %>%
  select(any_of(var))
```

#### Conditional select - AND, OR, NEG
```{r}
data %>%
  select(where(is.numeric)) %>%
  select(where(function(x) sd(x, na.rm = TRUE) > 2 & mean(x, na.rm = TRUE) < 10))
```
```{r}
data %>%
  select(where(is.numeric)) %>%
  select(where(function(x) sd(x, na.rm = TRUE) > 2 | mean(x, na.rm = TRUE) < 10 | median(x, na.rm = TRUE) > 5))
```
```{r}
data %>%
  select(!where(is.numeric))
```
#### Select other columns 
```{r}
data %>% 
  select(`Пол`, `Эритроциты_E1`, everything())
```
#### Select and change names
```{r}
data %>%
  select("Erythrocytes_V1" = `Эритроциты_E1`, "Erythrocytes_V2" = `Эритроциты_E2`)
```
### Select raws
```{r}
data %>%
  slice(1:10)
```
```{r}
data %>%
  slice_head(n = 10)
```
```{r}
data %>%
  slice_tail(prop = 0.2)
```
```{r}
data %>%
  slice_sample(prop = 0.1)
```
```{r}
data %>%
  slice_min(`Возраст`)
```

## Mutation 

```{r}
data %>%
  mutate(`Women_4th_blood_type` = ifelse(`Пол` == "Женский" & `Группа.крови` == "AB (IV)", "Y", "N")) %>%
  select(`Women_4th_blood_type`, everything()) %>%
  arrange(desc(`Women_4th_blood_type`))
```

### Arithmetic operations
```{r}
tibble(var_1 = 1:10, var_2 = var_1 + 1.123) %>%
  mutate (var_sum = var_1 + var_2,
          var_minus = var_1 - var_2,
          var_multiple = var_1 * var_2,
          var_divide = var_1 / var_2,
          var_1_log = log(var_1),
          var_1_log1p = log1p(var_1),
          var_1_exp = exp(var_1_log),
          var_1_exm1 = expm1 (var_1_log1p),
          var_2_round = round (var_2, 2),
          var_2_ceil = ceiling (var_2),
          var_2_floor = floor (var_2)) %>%
  glimpse()
```

### Use a case-when option
```{r}
data %>%
  mutate(`Age_group` = case_when(`Возраст` < 20 ~ "<20",
                                 between(`Возраст`, 20, 30) ~ "20-30",
                                 `Возраст` > 30 ~ ">30") %>% as.factor()) %>%
  select(`Возраст`, `Age_group`) %>%
  arrange(`Возраст`)
```

### Replace NA
```{r}
data %>%
  mutate(`Группа.крови` =`Группа.крови` %>% as.character() %>% replace_na("No data") %>% as.factor())
```
```{r}
data %>%
  mutate(`Группа.крови` =`Группа.крови` %>% na_if("AB (IV)"))
```
### Delete variable through mutate()
```{r}
data %>% 
  mutate(`Группа.крови` = NULL)
```

### Across() function
```{r}
data %>%
  mutate(across(where(is.numeric), function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))
```

### Iteration through rows
```{r}
data %>%
  rowwise() %>%
  mutate(`Basophils_Mean` = mean(c_across(contains("Базофилы")))) %>%
  ungroup() %>%
  select(`Basophils_Mean`, contains("Базофилы"))
```
```{r}
data %>%
  group_by(`Группа`) %>%
  mutate(across(contains("Базофилы"), function(x) x - mean(x, na.rm = TRUE))) %>%
  ungroup() %>%
  select(`Группа`, contains("Базофилы"))
```

## Renaming
```{r}
data %>%
  rename(`Group` = `Группа`) %>%
  select(`Group`)
```
```{r}
data %>%
  rename_with(function(x) x %>% stri_replace_all_regex(c("_E1", "_E2"), c("_V1", "_V2"), vectorize_all =  FALSE)) %>%
  glimpse()
```
```{r}
new_name <- "Group"
data %>%
  rename(!! new_name := `Группа`) %>%
  glimpse()
```

## Sorting
```{r}
data %>%
  arrange(`Пол`, desc(`Возраст`))
```
```{r}
data %>%
  group_by(`Группа`) %>%
  arrange(`Возраст`, .by_group = TRUE)
```

## Pivoting 
```{r}
data %>%
  select(`Группа`, contains("_E1")) %>%
  pivot_longer(!`Группа`)
```
```{r}
data %>%
  select(`Группа`, contains("_E1")) %>%
  mutate(ID = row_number()) %>%
  pivot_longer(!c(`Группа`, ID)) %>%
  pivot_wider(id_cols = ID)
```

## Unique values choise
```{r}
data %>%
  distinct(`Группа`, .keep_all = TRUE)
```

## Separation and union
```{r}
tibble(var_1 = rep(paste0("first part", "__", "second_part"), 10)) %>% 
  separate(var_1, into = c("var_1", "var_2"), sep = "__")
```

```{r}
tibble(var_1 = rep(paste0("first part", "__", "second_part"), 10)) %>% 
  separate(var_1, into = c("var_1", "var_2"), sep = "__") %>%
  unite("new_var", var_1, var_2, sep = " AND ")
```
## Create statistical tables
```{r}
data <- read.csv("data/raw/data_csv.csv", stringsAsFactors = TRUE)
data %>%
  select(`Группа`, where(is.numeric)) %>%
  group_by(`Группа`) %>%
  summarise(across(where(is.numeric), function(x) mean(x, na.rm = TRUE)))
```

### Table for numerical variables
```{r}
statistics <- list(
  `_Количество субъектов` = ~as.character(length(.x)),
  `_Количество (есть данные)` = ~as.character(sum(!is.na(.x))),
  `_Нет данных` = ~as.character(sum(is.na(.x))),
  `_Ср. знач.` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `_Станд. отклон.` = ~ifelse(sum(!is.na(.x)) < 3, "Н/П*", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_95% ДИ для среднего` = ~ifelse(sum(!is.na(.x)) < 3, "Н/П*", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_мин. - макс.` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", paste0(as.character(min(.x, na.rm = TRUE) %>% round(2)), " - ", as.character(max(.x, na.rm = TRUE) %>% round(2)))),
  `_Медиана` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `_Q1 - Q3` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", paste0(as.character(quantile(.x, 0.25, na.rm = TRUE) %>% round(2)), " - ", as.character(quantile(.x, 0.75, na.rm = TRUE) %>% round(2))))
)
data %>%
  select(`Группа`, where(is.numeric)) %>%
  group_by(`Группа`) %>%
  summarise(across(where(is.numeric), statistics)) %>%
  pivot_longer(!`Группа`, names_sep = "__", names_to = c("Переменная", "Статистика")) %>%
  rename(`Значение` = value)
```

### Table for categorical variables
```{r}
data %>%
  select(`Группа`, where(is.factor)) %>%
  mutate(`Группа.крови` = `Группа.крови` %>% as.character() %>% replace_na("Нет данных") %>% as.factor()) %>%
  count(`Группа`, `Группа.крови`) %>%
  group_by(`Группа`) %>%
  mutate(`Процент по группе` = (n / sum(n)) %>% round(4) %>% `*` (100) %>% str_c("%")) %>%
  ungroup() %>%
  mutate(`Процент по выборке` = (n / sum(n)) %>% round(4) %>% `*` (100) %>% str_c("%")) %>%
```

### FlexTable

```{r}
statistics <- list(
  `_Количество субъектов` = ~as.character(length(.x)),
  `_Количество (есть данные)` = ~as.character(sum(!is.na(.x))),
  `_Нет данных` = ~as.character(sum(is.na(.x))),
  `_Ср. знач.` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `_Станд. отклон.` = ~ifelse(sum(!is.na(.x)) < 3, "Н/П*", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_95% ДИ для среднего` = ~ifelse(sum(!is.na(.x)) < 3, "Н/П*", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_мин. - макс.` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", paste0(as.character(min(.x, na.rm = TRUE) %>% round(2)), " - ", as.character(max(.x, na.rm = TRUE) %>% round(2)))),
  `_Медиана` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `_Q1 - Q3` = ~ifelse(sum(!is.na(.x)) == 0, "Н/П*", paste0(as.character(quantile(.x, 0.25, na.rm = TRUE) %>% round(2)), " - ", as.character(quantile(.x, 0.75, na.rm = TRUE) %>% round(2))))
)
data %>%
  select(`Группа`, where(is.numeric)) %>%
  group_by(`Группа`) %>%
  summarise(across(where(is.numeric), statistics)) %>%
  pivot_longer(!`Группа`, names_sep = "__", names_to = c("Переменная", "Статистика")) %>%
  rename(`Значение` = value) %>%
  flextable() %>%
  theme_zebra() %>%
  merge_v(c("Группа", "Переменная"))
```
#### Merge
```{r}
tibble(var_1 = c("p-value", "0.001"), var_2 = c("p-value", "0.05")) %>%
      flextable() %>%
      theme_box() %>%
      merge_h(i = 1)
```
#### Align
```{r}
tibble(`Adverse events` = c("SOC Желудочно-кишечные нарушения 10017947", "PT Тошнота 10028813", "SOC Нарушения со стороны нервной системы 10029205", "PT Головная боль 10019211")) %>%
      flextable() %>%
      theme_box() %>%
      align(align = "center", part = "all")
```

#### Bold and italic
```{r}
tibble(`Adverse events` = c("SOC Желудочно-кишечные нарушения 10017947", "PT Тошнота 10028813", "SOC Нарушения со стороны нервной системы 10029205", "PT Головная боль 10019211")) %>%
      flextable() %>%
      theme_box() %>%
      align(i =~ str_detect(`Adverse events`, "SOC"), align = "left") %>%
      align(i =~ str_detect(`Adverse events`, "PT"), align = "right") %>%
      
      bold(i =~ str_detect(`Adverse events`, "SOC")) %>%
      italic(i =~ str_detect(`Adverse events`, "PT")) %>%
  
      width(width = 2)
```
#### Coloring
```{r}
is_pvalue <- function(x) {
  x %>% str_remove("<") %>% as.numeric() %>% `<`(0.05)
}

tibble("p-value" = c("<0.001", "0.38", "0.124", "0.005", "0.05")) %>%
  flextable() %>%
  theme_box() %>%
  color(i =~ is_pvalue(`p-value`), color = "green")
```
```{r}
is_pvalue <- function(x) {
  x %>% str_remove("<") %>% as.numeric() %>% `<`(0.05)
}

tibble("p-value" = c("<0.001", "0.38", "0.124", "0.005", "0.05")) %>%
  flextable() %>%
  theme_box() %>%
  bg(i =~ is_pvalue(`p-value`), bg = "green")
```
