---
title: "Study of the predictive ability of immunological status and clustering features on the development of chronic graft-versus-host disease after allogeneic hematopoietic stem cell transplantation."
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r libraries}
library(dplyr)
library(stringr)
library(tidyr)
library(flextable)
library(officer)
library(tibble)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(caret)
library(corrplot)
library(randomForestSRC)
library(survival)
library(Boruta)
```

# 1. Read and Clean Data

```{r clean data}
# Read data and rename columns
df <- read.csv("../../raw_data/AI_Tcells_для ЕА - Sheet1.csv")

# Check duplicates
duplicates <- df %>%
  group_by(ID, Time_OS, cGVHD_time, Names) %>%
  filter(n() > 1) %>%
  ungroup()  

# Clean the data 
df_clean <- df %>%
  mutate(
    Abs_Value = as.numeric(str_replace_all(Abs_Value, ",", ".")),
    Abs_Value = as.numeric(Abs_Value)) %>%
  # Rename columns
  rename(
    Subject_ID = ID,
    Observation_Days_Count = Time_OS,
    cGVHD_Diagnosis_Day = cGVHD_time,
    Cell_Count = Abs_Value
  ) %>%
  mutate(
    # Add flag to indicate which patients experienced cGVHD
    cGVHD_flag = as.numeric(ifelse(!is.na(cGVHD_Diagnosis_Day), 1, 0)),
    Subject_ID = as.factor(Subject_ID),
    
    # Add blood test group 
    Blood_test_day = case_when(
      grepl('ДЕНЬ ХРРТПХ_ПЕР.КРОВЬ', df$Names) ~ "cGVHD",
      grepl('ДЕНЬ ХРРТПХ +', df$Names) ~ paste0(str_extract(df$Names, "(?<=\\+)\\d+(?=_)"), "_cGVHD"),
      TRUE ~ str_extract(df$Names, "(?<=\\+)\\d+(?=_)")
    ),
    
    # Create exact cell name
    Cell_name = str_extract(Names, "(?<=/).*")
  ) %>%
  select(-Names)

unify_cell_name <- function(cell_name, replacements) {
  cell_name_unified <- cell_name
  for (replacement in replacements) {
    old_value <- replacement[1]
    new_value <- replacement[2]
    
    old_value <- str_replace_all(old_value, "([\\+\\*\\?\\|\\(\\)\\[\\]\\{\\}\\^\\$\\.])", "\\\\\\1")
    cell_name_unified <- str_replace_all(cell_name_unified, old_value, new_value)
  }
  return(cell_name_unified)
}

replacements <- list(
  c("PD1", "PD-1"),
  c("СТАР2", "STAR2"),
  c("4", "CD4_"),
  c("CD4_+", "CD4+_"),
  c("8", "CD8_"),
  c("CD8_+", "CD8+_"),
  c("Th", "TH"),
  c("__", "_"),
  c("_ ", "_")
)

# Apply the function to dataframe
df_clean <- df_clean %>%
  mutate(
    Cell_name_unified = Cell_name,
    Cell_name_unified = unify_cell_name(Cell_name, replacements)
    )

# Check for duplicates for control 
duplicates <- df_clean %>%
  group_by(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, Blood_test_day, Cell_name) %>%
  filter(n() > 1) %>%
  ungroup() 

# Check the unique cells name 
unique_cells_name <- unique(df_clean$Cell_name_unified)
print(unique_cells_name)

rm(duplicates, replacements, unify_cell_name)
```

**Crucial steps:**

1.  cGVHD Flag: Adds a flag (cGVHD_flag) to indicate if a patient experienced chronic GVHD (1 if cGVHD_Diagnosis_Day is not NA, otherwise 0).
2.  Blood Test Timing: classifies the blood test timing into groups (e.g., cGVHD, +30_cGVHD).
3.  Cell Name Parsing: Extracts the exact immune cell name from the Names column.
4.  Defines a function unify_cell_name to standardize cell names using a list of replacements (e.g., PD1 -\> PD-1, 4 -\> CD4\_).

## 1.1. Rearrange Data

```{r rearrange data}

# Check the number of unique days
unique_days <- unique(df_clean$Blood_test_day)
print(unique_days)

# Transform the dataframe
df_transformed <- df_clean %>%
  group_by(Subject_ID, Blood_test_day) %>%
  pivot_wider(
    id_cols = c(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, Blood_test_day),
    names_from = Cell_name_unified,
    values_from = Cell_Count,
    values_fill = list(Cell_Count = NA)
  ) %>%
  ungroup()

# Check if in any cell contain more than 1 number
multi_value_cells <- apply(df_transformed, 2, function(column) {
  any(grepl(",", column, fixed = TRUE) | grepl(" ", column))
})

contains_multiple_values <- ifelse(any(multi_value_cells), "Yes", "No")
print(contains_multiple_values)

rm(unique_days, contains_multiple_values)
```

**Crucial steps:**

1.  Reshape the data, creating separate columns for each combination of Cell_name_unified and Blood_test_day. Missing values are filled with NA.
2.  Checks if any transformed cell contains multiple values
3.  Randomly selects a patient and compares their raw Cell_Count data.

# 2. Prediction of cGVHD outcome

## 2.1. Define DataSet

Survival Analysis are planned.

*Event:* Diagnosis of cGVHD (cGVHD_flag == 1).

*Survival Time:*

-   For patients with cGVHD: cGVHD_Diagnosis_Day as the survival time.
-   For patients without cGVHD: Observation_Days_Count as the censored survival time.

*Perform PCA to reduce 167 variables into fewer components that explain the maximum variance!*

```{r define dataset Prediction of cGVHD outcome}

# Select days of interest
df_cGVHD_prediction <- df_transformed %>%
  filter(Blood_test_day %in% c("90", "180", "365", "cGVHD")) 


```

## 2.2. Check for Missing Data

```{r check for missing data}
# Calculate missing values by Blood_test_day
missing_summary <- df_cGVHD_prediction %>%
  group_by(Blood_test_day) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(cols = -Blood_test_day, names_to = "Column", values_to = "Missing_Percentage")
```

**Crucial steps:**

1.  No missing values by Blood test day were identified for any cell name.

## 2.3. Visualisation

### 2.3.1. Correlation martrix

```{r correlation matrix}
# Select relevant numeric columns 
correlation_data <- df_cGVHD_prediction %>%
  filter(Blood_test_day %in% c("90")) %>%
  filter(cGVHD_Diagnosis_Day > 90 | is.na(cGVHD_Diagnosis_Day)) %>%
  select(where(is.numeric))

cor_matrix <- cor(correlation_data, use = "complete.obs")

# Mask weak correlations
cor_matrix[abs(cor_matrix) < 0.5] <- NA

# Plot strong correlations
corrplot(cor_matrix, 
         method = "color", 
         type = "lower", 
         tl.col = "black", 
         tl.srt = 45, 
         number.cex = 0.1, 
         addCoef.col = "black", 
         mar = c(2, 2, 2, 2),
         cl.cex = 0.5,        
         tl.cex = 0.15)

correlation_table <- as.data.frame(as.table(cor_matrix))

colnames(correlation_table) <- c("Variable1", "Variable2", "Correlation")

# Filter for correlations >= 0.9
filtered_table <- correlation_table %>%
  filter(Correlation >= 0.9 & Variable1 != Variable2)

# Ensure no duplicate pairs
filtered_table <- filtered_table %>%
  mutate(
    Pair = paste0(pmin(Variable1, Variable2), "-", pmax(Variable1, Variable2))) %>%
  distinct(Pair, .keep_all = TRUE) %>%
  select(-Pair) %>%
  arrange(desc(Correlation)) %>%
  mutate(Num = row_number())

print(filtered_table)

# Check correlations for 'cGVHD_Diagnosis_Day' >= 0.6
cgvhd_filtered_table <- correlation_table %>%
  filter(
    (Variable1 == "cGVHD_Diagnosis_Day") &
    Correlation > 0.6 & 
    (Variable1 != Variable2)
  ) %>%
  arrange(desc(Correlation)) %>%
  mutate(Num = row_number())

# Print the filtered table
print(cgvhd_filtered_table)

```

-   In linear regression or logistic regression, multicollinearity makes it difficult to determine the unique contribution of each predictor variable.

Models that are not sensitive to multicollinearity because they split the data hierarchically:

-   *Random Forest:*
-   *XGBoost*

### 2.3.2. Heatmap with clusters

```{r visualisation heatmap with clusters all data}

correlation_data <- df_cGVHD_prediction %>%
  filter(Blood_test_day %in% c("90")) %>%
  filter(cGVHD_Diagnosis_Day > 90 | is.na(cGVHD_Diagnosis_Day)) %>%
  select(where(is.numeric), -cGVHD_Diagnosis_Day)

cor_matrix <- cor(correlation_data, use = "complete.obs")

# Hierarchical clustering and heatmap
pheatmap(cor_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = FALSE,
         fontsize_row = 2,
         fontsize_col = 2,
         main = "Clustered Heatmap for Survival Variables",
         width = 20, 
        height = 20  )


# Perform hierarchical clustering
distance_matrix <- as.dist(1 - cor_matrix) 
hclust_result <- hclust(distance_matrix, method = "ward.D2") # Ward's method

# Cut the dendrogram into clusters
clusters <- cutree(hclust_result, k = 6)

clustered_data <- data.frame(
  Variable = rownames(cor_matrix),
  Cluster = clusters
)

# Dendrogram with clusters
plot(hclust_result, labels = rownames(cor_matrix), main = "Dendrogram of Clusters", cex = 0.2)

# Add rectangles
rect.hclust(hclust_result, k = 6, border = 2:6)

```

```{r }
custom_order <- c("90", "180", "365", "cGVHD")

# Convert Blood_test_day to a factor with the custom order
df_cGVHD_transformed <- df_cGVHD_prediction %>%
  mutate(Blood_test_day = factor(Blood_test_day, levels = custom_order))

# Arrange data by the specified order of Blood_test_day
df_cGVHD_transformed <- df_cGVHD_transformed %>%
  arrange(Blood_test_day)

# Aggregate data by Blood_test_day
heatmap_data <- df_cGVHD_transformed %>%
  group_by(Blood_test_day) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  ungroup()

# Convert to matrix for heatmap plotting
heatmap_matrix <- heatmap_data %>%
  column_to_rownames("Blood_test_day") %>%
  as.matrix()

# Create the heatmap
pheatmap(heatmap_matrix,
         cluster_rows = FALSE,
         cluster_cols = TRUE,
         display_numbers = FALSE,
         fontsize_row = 6,
         fontsize_col = 3,
         main = "Heatmap: Variable Distribution by Blood Test Day")

```

Some variable distributions seem to change over time (Blood Test Day), with certain groups (e.g., those in warmer tones) showing higher activity at specific days, such as 365 (for non-log-transformed data).

### 2.3.3. PCA

```{r PCA}

# Filter and normalize data while preserving cGVHD_flag
df_normalized <- df_transformed %>%
  filter(Blood_test_day %in% c("90")) %>%
  filter(cGVHD_Diagnosis_Day > 90 | is.na(cGVHD_Diagnosis_Day)) %>%
  # Ensure cGVHD_flag is retained for later use
  select(cGVHD_flag, where(is.numeric), -Observation_Days_Count, -cGVHD_Diagnosis_Day) %>%
  # Apply log transformation only to numeric columns excluding cGVHD_flag
  mutate(across(-cGVHD_flag, ~ log(. + 1))) %>%
  # Scale numeric columns (excluding cGVHD_flag) to have zero mean and unit variance
  mutate(across(starts_with("log_"), ~ (.-mean(.)) / sd(.)))

# Save cGVHD_flag for later use
cGVHD_flags <- df_normalized$cGVHD_flag

# Perform near-zero variance filtering (excluding cGVHD_flag)
nzv <- nearZeroVar(df_normalized %>% select(-cGVHD_flag))
df_filtered <- df_normalized %>%
  select(-nzv, cGVHD_flag) %>% # Retain cGVHD_flag after filtering
  filter(complete.cases(.))

# Perform PCA (exclude cGVHD_flag from PCA computation)
pca_results <- prcomp(df_filtered %>% select(-cGVHD_flag), scale. = TRUE)

# Attach PCA scores and cGVHD_flag back together
pca_scores <- as.data.frame(pca_results$x) %>%
  mutate(cGVHD_flag = cGVHD_flags)

# Explained variance calculations
explained_variance <- pca_results$sdev^2 / sum(pca_results$sdev^2)
cumulative_variance <- cumsum(explained_variance)

# Create a data frame for plotting explained variance
explained_variance_data <- data.frame(
  PC = 1:length(explained_variance),
  Explained_Variance = explained_variance,
  Cumulative_Variance = cumulative_variance
)

# Filter for components contributing to the first 80% of cumulative variance
explained_variance_data_filtered <- explained_variance_data %>%
  filter(Cumulative_Variance <= 0.8)

# Add percentage labels for explained variance
explained_variance_data_filtered <- explained_variance_data_filtered %>%
  mutate(Variance_Percentage_Label = paste0(round(Explained_Variance * 100, 2), "%"))

# Plot explained variance with percentage labels
ggplot(explained_variance_data_filtered, aes(x = PC)) +
  geom_bar(aes(y = Explained_Variance), stat = "identity", fill = "steelblue") +
  geom_text(aes(y = Explained_Variance, label = Variance_Percentage_Label), 
            vjust = -0.5, size = 3.5) +
  geom_line(aes(y = Cumulative_Variance), color = "red", size = 1) +
  geom_point(aes(y = Cumulative_Variance), color = "red", size = 2) +
  scale_y_continuous(
    name = "Variance Explained",
    sec.axis = sec_axis(~., name = "Cumulative Variance Explained")
  ) +
  labs(
    title = "Explained Variance by Principal Components (First 80%)",
    x = "Principal Component"
  ) +
  theme_minimal(base_size = 14)

# Perform clustering on PCA scores
set.seed(123)
wss <- sapply(1:10, function(k) {
  kmeans(pca_scores[, 1:10], centers = k, nstart = 25)$tot.withinss
})

# Plot Elbow Method
elbow_plot <- data.frame(Clusters = 1:10, WSS = wss)
ggplot(elbow_plot, aes(x = Clusters, y = WSS)) +
  geom_line() +
  geom_point(size = 3) +
  labs(
    title = "Elbow Method for Optimal Clusters",
    x = "Number of Clusters",
    y = "Total Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 14)

# Apply K-means clustering
optimal_clusters <- 4
kmeans_result <- kmeans(pca_scores[, 1:10], centers = optimal_clusters, nstart = 25)
pca_scores$Cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters with coloring by cGVHD_flag
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(cGVHD_flag))) +
  geom_point(size = 2, alpha = 0.8) +  # Scatterplot of PCA points
  stat_ellipse(aes(group = Cluster), type = "t", linetype = "dashed", size = 0.5, color = "black") +  # Ellipses for clusters
  scale_color_brewer(palette = "Set1") +  
  labs(
    title = "PCA Clusters with cGVHD flag",
    x = "PC1 (Principal Component 1)",
    y = "PC2 (Principal Component 2)",
    color = "cGVHD_flag"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )


```

### 2.3.4. MDS

```{r MDS}

# Compute distance matrix (Euclidean distance in this case)
distance_matrix <- dist(df_filtered %>% select(-cGVHD_flag))

# Perform Multidimensional Scaling (MDS)
mds_results <- cmdscale(distance_matrix, k = 2) # k = 2 for 2D projection

# Convert MDS results to a data frame
mds_df <- as.data.frame(mds_results)
colnames(mds_df) <- c("MDS1", "MDS2")

# Add cluster information and cGVHD_flag to the MDS data
mds_df$Cluster <- as.factor(kmeans_result$cluster)
mds_df$cGVHD_flag <- df_filtered$cGVHD_flag

# Ensure Cluster is a factor
# Ensure Cluster and cGVHD_flag are factors
mds_df$Cluster <- as.factor(mds_df$Cluster)
mds_df$cGVHD_flag <- as.factor(mds_df$cGVHD_flag)

# Plot the MDS results
# Scale the y-axis by focusing on a tighter range or applying a log transformation
ggplot(mds_df, aes(x = MDS1, y = MDS2, shape = Cluster, color = cGVHD_flag)) +
  geom_point(size = 3, alpha = 0.8, position = position_jitter(width = 0.2, height = 0.2)) +
  scale_color_manual(values = c("red", "blue")) +
  stat_ellipse(aes(group = Cluster), type = "norm", linetype = "dashed", size = 0.5) +
  labs(
    title = "MDS Plot with Clusters and cGVHD flag",
    x = "MDS Dimension 1",
    y = "MDS Dimension 2",
    color = "cGVHD flag",
    shape = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "gray", size = 0.5, linetype = "dotted")
  )

```

Points belonging to different clusters (represented by shapes) are overlapping, particularly in the region close to the origin, indicating proximity in the MDS dimensions. Some points are further apart from the main cluster, such as the red circular points near the top-left, which could represent potential outliers. Flag-Based Separation:

The color-coded cGVHD_flag (red for 0, blue for 1) shows some separation within clusters, with a mix of both flags in the same region, implying limited distinction between flags in these dimensions.

#### 2.3.4.1. isoMDS

```{r isoMDS}
# Perform isoMDS (Non-metric MDS)
iso_mds_results <- MASS::isoMDS(distance_matrix, k = 2) # k = 2 for 2D projection

# Convert isoMDS results to a data frame
iso_mds_df <- as.data.frame(iso_mds_results$points)
colnames(iso_mds_df) <- c("MDS1", "MDS2")

# Add cluster information and cGVHD_flag
iso_mds_df$Cluster <- as.factor(kmeans_result$cluster)
iso_mds_df$cGVHD_flag <- df_filtered$cGVHD_flag

# Convert cGVHD_flag to a factor
iso_mds_df$cGVHD_flag <- as.factor(iso_mds_df$cGVHD_flag)

# Filter out clusters with fewer than 3 points (needed for ellipse calculation)
iso_mds_df_filtered <- iso_mds_df %>%
  group_by(Cluster) %>%
  filter(n() >= 3) %>%
  ungroup()

# Plot the isoMDS results with ellipses and adjusted axis scales
ggplot(iso_mds_df_filtered, aes(x = MDS1, y = MDS2, shape = Cluster, color = cGVHD_flag)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("red", "blue")) + # Define colors for cGVHD_flag
  stat_ellipse(aes(group = Cluster), type = "norm", linetype = "dashed", size = 0.5) + # Add ellipses
  labs(
    title = "isoMDS Plot with Clusters and cGVHD flag",
    x = "isoMDS Dimension 1",
    y = "isoMDS Dimension 2",
    color = "cGVHD_flag",
    shape = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "gray", size = 0.5, linetype = "dotted")
  )
```
