---
title: "Study of the predictive ability of immunological status and clustering features on the development of chronic graft-versus-host disease after allogeneic hematopoietic stem cell transplantation."
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r libraries}
library(dplyr)
library(stringr)
library(tidyr)
library(flextable)
library(officer)
library(tibble)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(caret)
library(corrplot)
library(randomForestSRC)
library(survival)
library(Boruta)
library(survAUC)
library(pROC)
```

# 1. Read and Clean Data

``` {r clean data}
# Read data and rename columns
df <- read.csv("../../raw_data/AI_Tcells_для ЕА - Sheet1.csv")

# Check duplicates
duplicates <- df %>%
  group_by(ID, Time_OS, cGVHD_time, Names) %>%
  filter(n() > 1) %>%
  ungroup()  

# Clean the data 
df_clean <- df %>%
  mutate(
    Abs_Value = as.numeric(str_replace_all(Abs_Value, ",", ".")),
    Abs_Value = as.numeric(Abs_Value)) %>%
  # Rename columns
  rename(
    Subject_ID = ID,
    Observation_Days_Count = Time_OS,
    cGVHD_Diagnosis_Day = cGVHD_time,
    Cell_Count = Abs_Value
  ) %>%
  mutate(
    # Add flag to indicate which patients experienced cGVHD
    cGVHD_flag = as.numeric(ifelse(!is.na(cGVHD_Diagnosis_Day), 1, 0)),
    Subject_ID = as.factor(Subject_ID),
    
    # Add blood test group 
    Blood_test_day = case_when(
      grepl('ДЕНЬ ХРРТПХ_ПЕР.КРОВЬ', df$Names) ~ "cGVHD",
      grepl('ДЕНЬ ХРРТПХ +', df$Names) ~ paste0(str_extract(df$Names, "(?<=\\+)\\d+(?=_)"), "_cGVHD"),
      TRUE ~ str_extract(df$Names, "(?<=\\+)\\d+(?=_)")
    ),
    
    # Create exact cell name
    Cell_name = str_extract(Names, "(?<=/).*")
  ) %>%
  select(-Names)

unify_cell_name <- function(cell_name, replacements) {
  cell_name_unified <- cell_name
  for (replacement in replacements) {
    old_value <- replacement[1]
    new_value <- replacement[2]
    
    old_value <- str_replace_all(old_value, "([\\+\\*\\?\\|\\(\\)\\[\\]\\{\\}\\^\\$\\.])", "\\\\\\1")
    cell_name_unified <- str_replace_all(cell_name_unified, old_value, new_value)
  }
  return(cell_name_unified)
}

replacements <- list(
  c("PD1", "PD-1"),
  c("СТАР2", "STAR2"),
  c("4", "CD4_"),
  c("CD4_+", "CD4+_"),
  c("8", "CD8_"),
  c("CD8_+", "CD8+_"),
  c("Th", "TH"),
  c("__", "_"),
  c("_ ", "_")
)

# Apply the function to dataframe
df_clean <- df_clean %>%
  mutate(
    Cell_name_unified = Cell_name,
    Cell_name_unified = unify_cell_name(Cell_name, replacements)
    )

# Check for duplicates for control 
duplicates <- df_clean %>%
  group_by(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, Blood_test_day, Cell_name) %>%
  filter(n() > 1) %>%
  ungroup() 

# Check the unique cells name 
unique_cells_name <- unique(df_clean$Cell_name_unified)
print(unique_cells_name)

rm(duplicates, replacements, unify_cell_name)
```

**Crucial steps:**

1. cGVHD Flag: Adds a flag (cGVHD_flag) to indicate if a patient experienced chronic GVHD (1 if cGVHD_Diagnosis_Day is not NA, otherwise 0).
2. Blood Test Timing: classifies the blood test timing into groups (e.g., cGVHD, +30_cGVHD).
3. Cell Name Parsing: Extracts the exact immune cell name from the Names column.
4. Defines a function unify_cell_name to standardize cell names using a list of replacements (e.g., PD1 -> PD-1, 4 -> CD4_).

## 1.1. Rearrange Data

```{r rearrange data}

# Check the number of unique days
unique_days <- unique(df_clean$Blood_test_day)
print(unique_days)

# Transform the dataframe
df_transformed <- df_clean %>%
  group_by(Subject_ID, Blood_test_day) %>%
  pivot_wider(
    id_cols = c(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, Blood_test_day),
    names_from = Cell_name_unified,
    values_from = Cell_Count,
    values_fill = list(Cell_Count = NA)
  ) %>%
  ungroup()

# Check if in any cell contain more than 1 number
multi_value_cells <- apply(df_transformed, 2, function(column) {
  any(grepl(",", column, fixed = TRUE) | grepl(" ", column))
})

contains_multiple_values <- ifelse(any(multi_value_cells), "Yes", "No")
print(contains_multiple_values)

rm(unique_days, contains_multiple_values)
```
**Crucial steps:**

1. Reshape the data, creating separate columns for each combination of Cell_name_unified and Blood_test_day. Missing values are filled with NA.
2. Checks if any transformed cell contains multiple values 
3. Randomly selects a patient and compares their raw Cell_Count data.

# 2. Prediction of cGVHD outcome
## 2.1. Define DataSet

Survival Analysis are planned. 

*Event:* Diagnosis of cGVHD (cGVHD_flag == 1).

*Survival Time:*

- For patients with cGVHD: cGVHD_Diagnosis_Day as the survival time.
- For patients without cGVHD: Observation_Days_Count as the censored survival time.

*Perform PCA to reduce 167 variables into fewer components that explain the maximum variance!*

## 2.2. Check for Missing Data

```{r check for missing data}
# Calculate missing values by Blood_test_day
missing_summary <- df_cGVHD_prediction %>%
  group_by(Blood_test_day) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(cols = -Blood_test_day, names_to = "Column", values_to = "Missing_Percentage")
```

**Crucial steps:**

1. No missing values by Blood test day were identified for any cell name.


### 2.3. BORUTA algorithm
#### 2.3.1.   90 days

```{r Boruta 90 days}
set.seed(123)

df_boruta_90 <- df_transformed %>%
  filter(Blood_test_day %in% c("90")) %>%
  filter(cGVHD_Diagnosis_Day > 90 | is.na(cGVHD_Diagnosis_Day)) %>%
  select(-Observation_Days_Count, -cGVHD_Diagnosis_Day, -Blood_test_day, -Subject_ID) %>%
  # Apply log transformation only to numeric columns excluding cGVHD_flag
  mutate(across(-cGVHD_flag, ~ log(. + 1))) %>%
  # Scale numeric columns (excluding cGVHD_flag) to have zero mean and unit variance
  mutate(across(starts_with("log_"), ~ (.-mean(.)) / sd(.)))

# Apply the Boruta algorithm
boruta_result_90 <- Boruta(cGVHD_flag ~ ., data = df_boruta_90)

# Extract important variables
important_vars_90 <- getSelectedAttributes(boruta_result_90, withTentative = FALSE)

# Extract tentative variables
tentative_vars_90 <- getSelectedAttributes(boruta_result_90, withTentative = TRUE)
only_tentative_90 <- setdiff(tentative_vars_90, important_vars_90)

cat("Confirmed important variables:\n", paste(important_vars_90, collapse = ", "), "\n\n")
cat("Tentative variables:\n", paste(only_tentative_90, collapse = ", "), "\n")

# keep only important and tentative features
filtered_importance <- boruta_result_90$ImpHistory[, tentative_vars_90, drop = FALSE]

# Filter only important and tentative features
filtered_important <- boruta_result_90$ImpHistory[, c(important_vars_90, only_tentative_90), drop = FALSE]

# Convert filtered_importance to a tidy format for ggplot
importance_data <- as.data.frame(filtered_importance) %>%
  pivot_longer(cols = everything(), names_to = "Feature", values_to = "Importance")

# Plot feature importance using ggplot
ggplot(importance_data, aes(x = reorder(Feature, Importance, FUN = median), y = Importance)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, outlier.color = "red", outlier.size = 1.5) +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Boruta Feature Importance (90 Days)",
    x = "Features",
    y = "Importance"
  ) +
  theme_minimal(base_size = 12) +  # Minimal theme with adjusted font size
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Center title
    axis.text.y = element_text(size = 8),  # Smaller text for y-axis labels
    axis.text.x = element_text(size = 10)
  )
```

```{r pca with boruta features}
# Keep only the selected important and tentative variables for PCA
df_pca_90 <- df_boruta_90 %>%
  select(all_of(c(important_vars_90, only_tentative_90)))

# Perform PCA
pca_result_90 <- prcomp(df_pca_90, center = TRUE, scale. = TRUE)

# Create a scree plot to show variance explained by each principal component
scree_data <- data.frame(PC = seq_along(pca_result_90$sdev),
                         Variance = (pca_result_90$sdev)^2 / sum((pca_result_90$sdev)^2))

ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_col(fill = "lightblue", alpha = 0.7) +
  geom_line(aes(group = 1), color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Scree Plot for PCA (90 Days)",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal(base_size = 12)

# Determine the optimal number of clusters using the Elbow Method
library(factoextra)
library(purrr)
kmeans_data <- pca_scores %>% select(PC1, PC2)
kmeans_wss <- map_dbl(1:10, function(k) {
  kmeans(kmeans_data, centers = k, nstart = 25)$tot.withinss
})

elbow_plot <- data.frame(Clusters = 1:10, WSS = kmeans_wss)

ggplot(elbow_plot, aes(x = Clusters, y = WSS)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Elbow Method for Optimal Clusters",
    x = "Number of Clusters",
    y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 12)

# Determine the optimal number of clusters using the Elbow Method
library(factoextra)
library(purrr)
kmeans_data <- pca_result_90$x %>% as.data.frame() %>% select(PC1, PC2)
kmeans_wss <- map_dbl(1:10, function(k) {
  kmeans(kmeans_data, centers = k, nstart = 25)$tot.withinss
})

elbow_plot <- data.frame(Clusters = 1:10, WSS = kmeans_wss)

ggplot(elbow_plot, aes(x = Clusters, y = WSS)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Elbow Method for Optimal Clusters",
    x = "Number of Clusters",
    y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal(base_size = 12)

# Perform k-means clustering with 4 clusters
kmeans_result <- kmeans(kmeans_data, centers = 2, nstart = 25)
pca_scores <- as.data.frame(pca_result_90$x)
pca_scores$Cluster <- factor(kmeans_result$cluster)
pca_scores$cGVHD_flag <- df_boruta_90$cGVHD_flag

# Plot PCA scatter plot with clusters and coloring by flag
pca_cluster_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = cGVHD_flag), alpha = 0.7, size = 2) +
  stat_ellipse(aes(fill = Cluster), level = 0.95, alpha = 0.2, geom = "polygon") +
  labs(
    title = "PCA Scatter Plot with 4 Clusters and Flag Coloring (90 Days)",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "cGVHD Flag",
    fill = "Cluster"
  ) +
  theme_minimal(base_size = 12)

print(pca_cluster_plot)


```

### 2.4.    Random Forest over Survival
#### 2.4.1    90 days

```{r Random Forest over Survival 90 days}
df_cGVHD_prediction_survival_90 <- df_cGVHD_prediction %>%
  filter(Blood_test_day %in% c("90")) %>%
  filter(cGVHD_Diagnosis_Day > 90 | is.na(cGVHD_Diagnosis_Day)) %>%
  mutate(
    cGVHD_Diagnosis_Day = ifelse(is.na(cGVHD_Diagnosis_Day), Observation_Days_Count, cGVHD_Diagnosis_Day),
    cGVHD_flag = as.numeric(as.character(cGVHD_flag))
  ) %>%
  select(-Subject_ID, -Blood_test_day, -Observation_Days_Count) %>%
  # Apply log transformation only to numeric columns excluding cGVHD_flag
  mutate(across(-c(cGVHD_flag, cGVHD_Diagnosis_Day), ~ log(. + 1))) %>%
  # Scale numeric columns (excluding cGVHD_flag) to have zero mean and unit variance
  mutate(across(starts_with("log_"), ~ (.-mean(.)) / sd(.)))

```

```{r Random Forest over Survival 90 days}
tune_rsf_model <- function(data, formula, ntree_values = c(50, 100, 200, 300, 400, 500), 
                           mtry_values = NULL, nodesize_values = c(5, 10, 15, 20)) {
  
  # Define default mtry values if not provided
  if (is.null(mtry_values)) {
    mtry_values <- c(round(sqrt(ncol(data) - 2)), 
                     round((ncol(data) - 2) / 3))
  }
  
  best_cindex <- 0
  best_params <- list(ntree = NA, mtry = NA, nodesize = NA)  # Use NA instead of NULL

  # Loop through all combinations of hyperparameters
  for (ntree in ntree_values) {
    for (mtry in mtry_values) {
      for (nodesize in nodesize_values) {
        
        # Try tuning the model and catch errors
        tune_results <- tryCatch({
          tune.rfsrc(formula, data = data, 
                     mtry = mtry, nodesize = nodesize, 
                     ntree = ntree, doBest = FALSE)
        }, error = function(e) NULL)  # Return NULL if error occurs
        
        # Ensure tune_results is valid
        if (!is.null(tune_results) && !is.null(tune_results$err.rate)) {
          current_cindex <- 1 - tune_results$err.rate[length(tune_results$err.rate)]
          
          # Update best parameters if current C-index is better
          if (!is.na(current_cindex) && current_cindex > best_cindex) {
            best_cindex <- current_cindex
            best_params <- list(ntree = ntree, mtry = mtry, nodesize = nodesize)
          }
        }
      }
    }
  }

  # Ensure best_params are numeric, otherwise set to default values
  if (is.na(best_params$ntree)) best_params$ntree <- 500
  if (is.na(best_params$mtry)) best_params$mtry <- round(sqrt(ncol(data) - 2))
  if (is.na(best_params$nodesize)) best_params$nodesize <- 10

  return(best_params)
}


# Set a seed for reproducibility
set.seed(123)

# Tune hyperparameters for the full dataset
best_params <- tune_rsf_model(
  data = df_cGVHD_prediction_survival_90, 
  formula = Surv(cGVHD_Diagnosis_Day, cGVHD_flag) ~ .
)

# Print best parameters
print(best_params)

# Train the best model
RF_obj <- rfsrc(
  Surv(cGVHD_Diagnosis_Day, cGVHD_flag) ~ .,
  data = df_cGVHD_prediction_survival_90,
  ntree = best_params$ntree,
  mtry = best_params$mtry,
  nodesize = best_params$nodesize,
  importance = TRUE,
  membership = TRUE
)

# Print model summary
print(RF_obj)

# Check variable importance
var_importance <- sort(RF_obj$importance, decreasing = TRUE)

# Print top 10 important features
print(var_importance[1:20])

# Select top 20 features for model refinement
top_features <- names(var_importance)[1:20]
df_selected <- df_cGVHD_prediction_survival_90[, c(top_features, "cGVHD_Diagnosis_Day", "cGVHD_flag")]

# Tune RSF on the selected features
best_params_selected <- tune_rsf_model(
  data = df_selected, 
  formula = Surv(cGVHD_Diagnosis_Day, cGVHD_flag) ~ .
)

# Train refined RSF model with top features
RF_obj_selected <- rfsrc(
  Surv(cGVHD_Diagnosis_Day, cGVHD_flag) ~ .,
  data = df_selected,
  ntree = best_params_selected$ntree,
  mtry = best_params_selected$mtry,
  nodesize = best_params_selected$nodesize,
  importance = TRUE,
  membership = TRUE
)

# Print final refined model
print(RF_obj_selected)

```

```{r survival curve 90 days}
# Predict survival probabilities for a specific patient
newdata <- data.frame(lapply(1:ncol(RF_obj_selected$xvar), function(i) { median(RF_obj_selected$xvar[, i]) }))
colnames(newdata) <- RF_obj_selected$xvar.names

# Generate predictions
y_pred <- predict(RF_obj_selected, newdata = newdata)

# Plot predicted survival curve
plot(
  y_pred$time.interest, 
  y_pred$survival[1, ], 
  type = "l", 
  col = "blue", 
  lwd = 2, 
  xlab = "Time (Days)", 
  ylab = "Survival Probability",
  main = "Predicted Survival Curve"
)
grid()


```



```{r Brier score 90 days}
# Calculate the Brier score using the Kaplan-Meier censoring distribution
bs.km <- get.brier.survival(RF_obj_selected, cens.mode = "km")$brier.score

# Plot the Brier score
plot(
  bs.km, 
  type = "s", 
  col = 2, 
  ylab = "Brier Score", 
  xlab = "Time (Days)", 
  main = "Brier Score Over Time"
)

```

- < 0.1 - excellent
- <= 0.2 - superior
- <=0.3 - adequate

```{r VIMP 90 days}
# Extract VIMP
vimp <- RF_obj_selected$importance

vimp_df <- data.frame(
  Variable = names(vimp),
  VIMP = vimp
)

# Filter variables above the threshold
vimp_df_positive <- vimp_df 

# Enhanced VIMP plot
ggplot(vimp_df_positive, aes(x = reorder(Variable, VIMP), y = VIMP)) +
  geom_bar(stat = "identity", fill = "lightblue", alpha = 0.8, color = "black", width = 0.7) +  # Bar aesthetics
  geom_text(aes(label = round(VIMP, 2)), hjust = -0.2, size = 3) +  # Add data labels
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Top Variable Importance (VIMP) in RSF",
    x = "Variables",
    y = "VIMP"
  ) +
  theme_minimal(base_size = 12) +  # Use a clean minimal theme
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),  # Centered and bold title
    axis.text.y = element_text(size = 10),  # Bold y-axis labels for better clarity
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    axis.title = element_text(size = 12, face = "bold"),  # Bold axis titles
    panel.grid.major.y = element_blank(),  # Remove horizontal grid lines for a cleaner look
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    legend.position = "none"  # No legend required
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))  # Add some space above the bars


```

```{r VIMP vs Boruta 90 days}
# Compare with VIMP
vimp_boruta_comparison <- vimp_df_positive %>%
  mutate(
    Boruta_Status = case_when(
      Variable %in% important_vars_90 ~ "Confirmed Important",
      Variable %in% only_tentative_90 ~ "Tentative",
      TRUE ~ "Rejected"
    )
  )


# Plot Comparison
ggplot(vimp_boruta_comparison, aes(x = reorder(Variable, VIMP), y = VIMP, fill = Boruta_Status)) +
  geom_bar(stat = "identity", color = "black", width = 0.7, alpha = 0.9) +  # Bar aesthetics
  coord_flip() +  # Flip coordinates for readability
  scale_fill_manual(
    values = c("Confirmed Important" = "yellow", "Tentative" = "lightgreen", "Rejected" = "lightblue"),
    labels = c("Confirmed Important", "Rejected", "Tentative")
  ) +
  labs(
    title = "Comparison of VIMP and Boruta Status",
    x = "Variables",
    y = "Variable Importance (VIMP)",
    fill = "Boruta Status"
  ) +
  theme_minimal(base_size = 12) +  # Minimal clean theme
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),  # Bold, centered title
    axis.text.y = element_text(size = 8),  # Bold y-axis labels
    axis.text.x = element_text(size = 8),  # X-axis text size
    axis.title = element_text(size = 12, face = "bold"),  # Bold axis titles
    legend.position = "bottom",  # Move legend to the top
    legend.title = element_text(face = "bold"),  # Bold legend title
    panel.grid.major.y = element_blank(),  # Remove horizontal grid lines
    panel.grid.minor = element_blank()  # Remove minor grid lines
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))  # Add space above bars


# Print Final Comparison Table
print(vimp_boruta_comparison)

```


## 2.5.   NMF, Boruta and Random Forest over Survival comparison
### 2.5.1.  NMF and Boruta Comparison 
```{r NMF Boruta comparison}

# Add the NMF results as a vector
nmf_features <- c(
  "8EMTM+DR+", "8EMTM+PD-1+TIGIT-", "8EMTM+PD-1+TIGIT+", "8EMTM+PD-1-TIGIT-", 
  "8EMTM+39+", "4EM+TIGIT+", "4EM+PD1+TIGIT+", "4TE+PD-1", "4+DR+", 
  "8+39+", "8TE+39+", "8+PD-1+TIGIT-", "4TM_TH17", "4EM+39+", 
  "4+PD-1+", "4TE+PD-1+TIGIT", "4TE+39+", "4+39+", "TREG_TE", 
  "8+PD-1+TIGIT-",
  "4NV+226+", "4NV+PD-1-TIGIT-", "4NV(STAR2)", "4NV", "4NV_TH2", 
  "TREG_NV", "4NV_TH1", "8NV+PD-1-TIGIT-", "4NV+TIGIT+"
)

# Extract Boruta results
boruta_features <- c("4СМ_TH2", "4СМ_TH22", "8EMTM+DR+", "8EMTM+PD-1+", "8EMTM+PD-1+TIGIT+", "8EMTM+TIGIT+", "8+", "8_EMTM", "8+226+")

# Compare features
shared_features <- intersect(nmf_features, boruta_features)  # Overlapping features
unique_to_nmf <- setdiff(nmf_features, boruta_features)      # Features unique to NMF
unique_to_boruta <- setdiff(boruta_features, nmf_features)   # Features unique to Boruta

# Print results
cat("Shared Features:\n", shared_features, "\n\n")
cat("Features Unique to NMF:\n", unique_to_nmf, "\n\n")
cat("Features Unique to Boruta:\n", unique_to_boruta, "\n\n")

# Visualize with a Venn diagram
library(VennDiagram)
# Generate Venn diagram and include feature names in the overlap
grid.newpage() # Clear the plotting window
venn.plot <- venn.diagram(
  x = list(NMF = nmf_features, Boruta = boruta_features),
  category.names = c("NMF", "Boruta"),
  filename = NULL, # Do not save the diagram
  fill = c("skyblue", "pink"), # Colors for NMF and Boruta
  alpha = 0.5, # Transparency for circles
  lty = "dashed", # Dashed borders for circles
  cex = 1.5, # Adjust numbers' font size
  cat.cex = 1.2, # Adjust labels' font size
  main = "Feature Comparison: NMF vs Boruta (Confirmed + Tentative)",
  main.cex = 1.5, # Title font size
  force.unique = TRUE # Prevent duplicate elements
)

# Add the feature names inside the overlapping region
shared_label <- paste(shared_features, collapse = "\n") # Combine shared features into one string
grid.text(
  shared_label,
  x = 0.7, # Position at the center of the overlap
  y = 0.7, # Adjust as needed
  gp = gpar(fontsize = 10) # Font size for the labels
)

# Draw the diagram
grid.draw(venn.plot)

```

### 2.5.2.  NMF and RFS Comparison
```{r NMF RFS comparison}

# Extract Boruta results
rfs_features <- c(
  "TREG_TE", "8EMTM+TIGIT+", "4NV_TH17TO1", "8EMTM+PD-1+TIGIT+",
      "8EMTM+PD-1+", "8EMTM+DR+", "TREG+PD-1+",
      "8TREG", "8+", "8+DR+", "4+TIGIT+", "4СМ_TH22", 
  "4ТЕ+PD-1+TIGIT+",  "8+(IM STAT)", "8+226+", "8+PD-1-TIGIT-", 
  "4ТM_TH2", "8EMTM", "8СМ+PD-1+TIGIT-", "4СМ(STAR2)"
)

# Compare features
shared_features <- intersect(nmf_features, rfs_features)  # Overlapping features
unique_to_nmf <- setdiff(nmf_features, rfs_features)      # Features unique to NMF
unique_to_boruta <- setdiff(boruta_features, rfs_features)   # Features unique to Boruta

# Print results
cat("Shared Features:\n", shared_features, "\n\n")
cat("Features Unique to NMF:\n", unique_to_nmf, "\n\n")
cat("Features Unique to Boruta:\n", unique_to_boruta, "\n\n")

# Visualize with a Venn diagram
library(VennDiagram)
# Generate Venn diagram and include feature names in the overlap
grid.newpage() # Clear the plotting window
venn.plot <- venn.diagram(
  x = list(NMF = nmf_features, Boruta = rfs_features),
  category.names = c("NMF", "RFS"),
  filename = NULL, # Do not save the diagram
  fill = c("skyblue", "pink"), # Colors for NMF and Boruta
  alpha = 0.5, # Transparency for circles
  lty = "dashed", # Dashed borders for circles
  cex = 1.5, # Adjust numbers' font size
  cat.cex = 1.2, # Adjust labels' font size
  main = "Feature Comparison: NMF vs RFS",
  main.cex = 1.5, # Title font size
  force.unique = TRUE # Prevent duplicate elements
)

# Add the feature names inside the overlapping region
shared_label <- paste(shared_features, collapse = "\n") # Combine shared features into one string
grid.text(
  shared_label,
  x = 0.6, # Position at the center of the overlap
  y = 0.7, # Adjust as needed
  gp = gpar(fontsize = 10) # Font size for the labels
)

# Draw the diagram
grid.draw(venn.plot)

```

```{r RFS NMF comparison by cell populations}
# Define cell population groups
effector_memory_4 <- unique(c(
  grep("(4EM|4EMTM|4TM)", nmf_features, value = TRUE), # Features from NMF
  grep("(4EM|4EMTM|4TM)", boruta_features, value = TRUE)  # Features from RFS
))

effector_memory_8 <- unique(c(
  grep("(8EM|8EMTM|8TM)", nmf_features, value = TRUE), # Features from NMF
  grep("(8EM|8EMTM|8TM)", boruta_features, value = TRUE)  # Features from RFS
))

naive_4 <- unique(c(
  grep("4NV", nmf_features, value = TRUE),
  grep("4NV", boruta_features, value = TRUE)
))

naive_8 <- unique(c(
  grep("8NV", nmf_features, value = TRUE),
  grep("8NV", boruta_features, value = TRUE)
))

central_memory_4 <- unique(c(
  grep("4CM", nmf_features, value = TRUE),
  grep("4CM", boruta_features, value = TRUE)
))

central_memory_8 <- unique(c(
  grep("8CM", nmf_features, value = TRUE),
  grep("8CM", boruta_features, value = TRUE)
))

regulatory <- unique(c(
  grep("TREG", nmf_features, value = TRUE),
  grep("TREG", boruta_features, value = TRUE)
))

# Combine results into a list
cell_populations <- list(
  EM_CD4 = effector_memory_4,
  EM_CD8 = effector_memory_8,
  Naive_CD4 = naive_4,
  Naive_CD8 = naive_8,
  CM_CD4 = central_memory_4,
  CM_CD8 = central_memory_8,
  Regulatory = regulatory
)

# Check which NMF features belong to each population
nmf_by_population <- lapply(cell_populations, function(pop) {
  intersect(nmf_features, pop)
})

# Check which RFS features belong to each population
boruta_by_population <- lapply(cell_populations, function(pop) {
  intersect(boruta_features, pop)
})

# Print results
cat("NMF Features by Population:\n")
print(nmf_by_population)

cat("\nBoruta Features by Population:\n")
print(boruta_features)

# Count features for each population
nmf_counts <- sapply(cell_populations, function(pop) {
  sum(pop %in% nmf_features)
})

boruta_counts <- sapply(cell_populations, function(pop) {
  sum(pop %in% boruta_features)
})

# Combine counts into a data frame
population_counts <- data.frame(
  Population = names(cell_populations),
  NMF = nmf_counts,
  Boruta = boruta_counts
)

print(population_counts)

# Reshape data for plotting
population_counts_long <- pivot_longer(
  population_counts,
  cols = c("NMF", "Boruta"),
  names_to = "Method",
  values_to = "Count"
)

ggplot(population_counts_long, aes(x = Population, y = Count, fill = Method)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black", width = 0.7) +
  geom_text(
    aes(label = Count), 
    position = position_dodge(width = 0.7), 
    vjust = -0.5, size = 4
  ) + # Add data labels
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Feature Distribution by Cell Population",
    x = "Cell Population",
    y = "Number of Features",
    fill = "Method"
  ) +
  theme_minimal(base_size = 14) + # Adjust base font size
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16), 
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.position = "top", 
    legend.title = element_text(face = "bold") 
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) 

```

```{r RFS NMF comparison by cell populations}
# Define cell population groups
effector_memory_4 <- unique(c(
  grep("(4EM|4EMTM|4TM)", nmf_features, value = TRUE), # Features from NMF
  grep("(4EM|4EMTM|4TM)", rfs_features, value = TRUE)  # Features from RFS
))

effector_memory_8 <- unique(c(
  grep("(8EM|8EMTM|8TM)", nmf_features, value = TRUE), # Features from NMF
  grep("(8EM|8EMTM|8TM)", rfs_features, value = TRUE)  # Features from RFS
))

naive_4 <- unique(c(
  grep("4NV", nmf_features, value = TRUE),
  grep("4NV", rfs_features, value = TRUE)
))

naive_8 <- unique(c(
  grep("8NV", nmf_features, value = TRUE),
  grep("8NV", rfs_features, value = TRUE)
))

central_memory_4 <- unique(c(
  grep("4CM", nmf_features, value = TRUE),
  grep("4CM", rfs_features, value = TRUE)
))

central_memory_8 <- unique(c(
  grep("8CM", nmf_features, value = TRUE),
  grep("8CM", rfs_features, value = TRUE)
))

regulatory <- unique(c(
  grep("TREG", nmf_features, value = TRUE),
  grep("TREG", rfs_features, value = TRUE)
))

# Combine results into a list
cell_populations <- list(
  EM_CD4 = effector_memory_4,
  EM_CD8 = effector_memory_8,
  Naive_CD4 = naive_4,
  Naive_CD8 = naive_8,
  CM_CD4 = central_memory_4,
  CM_CD8 = central_memory_8,
  Regulatory = regulatory
)

# Check which NMF features belong to each population
nmf_by_population <- lapply(cell_populations, function(pop) {
  intersect(nmf_features, pop)
})

# Check which RFS features belong to each population
rfs_by_population <- lapply(cell_populations, function(pop) {
  intersect(rfs_features, pop)
})

# Print results
cat("NMF Features by Population:\n")
print(nmf_by_population)

cat("\nRFS Features by Population:\n")
print(rfs_by_population)

# Count features for each population
nmf_counts <- sapply(cell_populations, function(pop) {
  sum(pop %in% nmf_features)
})

rfs_counts <- sapply(cell_populations, function(pop) {
  sum(pop %in% rfs_features)
})

# Combine counts into a data frame
population_counts <- data.frame(
  Population = names(cell_populations),
  NMF = nmf_counts,
  RFS = rfs_counts
)

print(population_counts)

# Reshape data for plotting
population_counts_long <- pivot_longer(
  population_counts,
  cols = c("NMF", "RFS"),
  names_to = "Method",
  values_to = "Count"
)

ggplot(population_counts_long, aes(x = Population, y = Count, fill = Method)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black", width = 0.7) +
  geom_text(
    aes(label = Count), 
    position = position_dodge(width = 0.7), 
    vjust = -0.5, size = 4
  ) + # Add data labels
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Feature Distribution by Cell Population",
    x = "Cell Population",
    y = "Number of Features",
    fill = "Method"
  ) +
  theme_minimal(base_size = 14) + # Adjust base font size
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16), 
    axis.text.x = element_text(angle = 45, hjust = 1), 
    legend.position = "top", 
    legend.title = element_text(face = "bold") 
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) 

```





