---
title: "Study of the predictive ability of immunological status and clustering features on the development of chronic graft-versus-host disease after allogeneic hematopoietic stem cell transplantation."
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r libraries}
library(dplyr)
library(stringr)
library(tidyr)
library(flextable)
library(officer)
library(tibble)
library(ggplot2)
library(reshape2)
library(pheatmap)
```

# 1. Read and Clean Data

``` {r clean data}
# Read data and rename columns
df <- readxl::read_excel("./data/AI_Tcells_DB.xlsx") 

# Check duplicates
duplicates <- df %>%
  group_by(ID, Time_OS, cGVHD_time, Names) %>%
  filter(n() > 1) %>%
  ungroup()  

# Clean the data 
df_clean <- df %>%
  # Rename columns
  rename(
    Subject_ID = ID,
    Observation_Days_Count = Time_OS,
    cGVHD_Diagnosis_Day = cGVHD_time,
    Cell_Count = Abs_Value
  ) %>%
  mutate(
    # Add flag to indicate which patients experienced cGVHD
    cGVHD_flag = as.numeric(ifelse(!is.na(cGVHD_Diagnosis_Day), 1, 0)),
    
    # Add blood test group 
    Blood_test_day = case_when(
      grepl('ДЕНЬ ХРРТПХ_ПЕР.КРОВЬ', df$Names) ~ "cGVHD",
      grepl('ДЕНЬ ХРРТПХ +', df$Names) ~ paste0(str_extract(df$Names, "(?<=\\+)\\d+(?=_)"), "_cGVHD"),
      TRUE ~ str_extract(df$Names, "(?<=\\+)\\d+(?=_)")
    ),
    
    # Create exact cell name
    Cell_name = str_extract(Names, "(?<=/).*")
  ) %>%
  select(-Names)

unify_cell_name <- function(cell_name, replacements) {
  cell_name_unified <- cell_name
  for (replacement in replacements) {
    old_value <- replacement[1]
    new_value <- replacement[2]
    
    old_value <- str_replace_all(old_value, "([\\+\\*\\?\\|\\(\\)\\[\\]\\{\\}\\^\\$\\.])", "\\\\\\1")
    cell_name_unified <- str_replace_all(cell_name_unified, old_value, new_value)
  }
  return(cell_name_unified)
}

replacements <- list(
  c("PD1", "PD-1"),
  c("СТАР2", "STAR2"),
  c("4", "CD4_"),
  c("CD4_+", "CD4+_"),
  c("8", "CD8_"),
  c("CD8_+", "CD8+_"),
  c("Th", "TH"),
  c("__", "_"),
  c("_ ", "_")
)

# Apply the function to dataframe
df_clean <- df_clean %>%
  mutate(
    Cell_name_unified = Cell_name,
    Cell_name_unified = unify_cell_name(Cell_name, replacements)
    )

# Check for duplicates for control 
duplicates <- df_clean %>%
  group_by(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, Blood_test_day, Cell_name) %>%
  filter(n() > 1) %>%
  ungroup() 

# Check the unique cells name 
unique_cells_name <- unique(df_clean$Cell_name_unified)
print(unique_cells_name)

rm(duplicates, replacements, unify_cell_name)
```

**Crucial steps:**

1. cGVHD Flag: Adds a flag (cGVHD_flag) to indicate if a patient experienced chronic GVHD (1 if cGVHD_Diagnosis_Day is not NA, otherwise 0).
2. Blood Test Timing: classifies the blood test timing into groups (e.g., cGVHD, +30_cGVHD).
3. Cell Name Parsing: Extracts the exact immune cell name from the Names column.
4. Defines a function unify_cell_name to standardize cell names using a list of replacements (e.g., PD1 -> PD-1, 4 -> CD4_).

## 1.1. Rearrange Data

```{r rearrange data}

# Check the number of unique days
unique_days <- unique(df_clean$Blood_test_day)
print(unique_days)

# Transform the dataframe
df_transformed <- df_clean %>%
  mutate(new_column = paste(Cell_name_unified, Blood_test_day, sep = "_")) %>%
  pivot_wider(
    id_cols = c(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag),
    names_from = new_column, 
    values_from = Cell_Count, 
    values_fill = list(Cell_Count = NA)
  )

# Check if in any cell contain more than 1 number
multi_value_cells <- apply(df_transformed, 2, function(column) {
  any(grepl(",", column, fixed = TRUE) | grepl(" ", column))
})

contains_multiple_values <- ifelse(any(multi_value_cells), "Yes", "No")
print(contains_multiple_values)

set.seed(123)

# Select a random patient
random_patient <- sample(unique(df_clean$Subject_ID), 1)

df_clean_sorted <- df_clean %>%
  filter(Subject_ID == random_patient)
clean_vector <- df_clean_sorted$Cell_Count

transformed_vector <- df_transformed %>%
  filter(Subject_ID == random_patient) %>%
  select(-Subject_ID, -Observation_Days_Count, -cGVHD_Diagnosis_Day, -cGVHD_flag) %>%
  unlist() %>%
  as.numeric()

transformed_vector <- transformed_vector[!is.na(transformed_vector)]

if (length(clean_vector) != length(transformed_vector)) {
  print(paste("Mismatch in lengths for patient:", random_patient))
  print(paste("Length of clean vector:", length(clean_vector)))
  print(paste("Length of transformed vector:", length(transformed_vector)))
} else {
  mismatches <- which(clean_vector != transformed_vector)
  
  if (length(mismatches) == 0) {
    print(paste("All values match for patient:", random_patient))
  } else {
    print(paste("Mismatches found for patient:", random_patient))
    print("Indexes of mismatches:")
    print(mismatches)
    print("Clean vector values at mismatched indexes:")
    print(clean_vector[mismatches])
    print("Transformed vector values at mismatched indexes:")
    print(transformed_vector[mismatches])
  }
}

# Function to rearrange columns
rearrange_columns <- function(data) {
  # Replace "__" with "_" in column names
  colnames(data) <- gsub("__", "_", colnames(data))
  
  # Define the order 
  desired_order <- c("_30", "_60", "_90", "_180", "_365", "_cGVHD", "_30cGVHD", "_60cGVHD", "_90cGVHD", "_180cGVHD")
  fixed_columns <- c("Subject_ID", "Observation_Days_Count", "cGVHD_Diagnosis_Day", "cGVHD_flag")
  other_columns <- setdiff(colnames(data), fixed_columns)
  
  ordered_columns <- unlist(lapply(desired_order, function(suffix) {
    grep(paste0(suffix, "$"), other_columns, value = TRUE)
  }))
  
  # Combine fixed columns
  final_order <- c(fixed_columns, ordered_columns)
  
  data <- data[, final_order, drop = FALSE]
  
  return(data)
}

# Apply the function
df_transformed <- rearrange_columns(df_transformed)

rm(unique_days, df_clean_sorted, clean_vector, contains_multiple_values, mismatches, multi_value_cells, random_patient, transformed_vector)
```
**Crucial steps:**

1. Reshape the data, creating separate columns for each combination of Cell_name_unified and Blood_test_day. Missing values are filled with NA.
2. Checks if any transformed cell contains multiple values 
3. Randomly selects a patient and compares their raw Cell_Count data.

# 2. Descriptive Statistics

``` {r descriptive statistics}
# Prepare df_transformed
df_transformed <- df_transformed %>%
  mutate(
    Subject_ID = as.factor(Subject_ID),
    cGVHD_flag = as.factor(cGVHD_flag)
  )

# List of descriptive statistics
statistics <- list(
  `Number of values` = ~as.character(sum(!is.na(.x))),
  `No data` = ~as.character(sum(is.na(.x))),
  `Mean` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `Median` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `SD` = ~ifelse(sum(!is.na(.x)) < 3, "NA", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `min` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(min(.x, na.rm = TRUE) %>% round(2))),
  `max` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(max(.x, na.rm = TRUE) %>% round(2)))
)

# Summarize the statistics for each numeric variable and make variables into columns
df_summary <- df_transformed %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), statistics, .names = "{.col}__{.fn}")) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat"), values_to = "Value") %>%
  pivot_wider(names_from = Stat, values_from = Value) %>%
  slice(1:25)

# Create a flextable for the summary
flextable(df_summary) %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))

# Clean and summarize the categorical data for all factor variables
df_transformed %>%
  select(-Subject_ID) %>%
  select(where(is.factor)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))

rm(df_summary, statistics)
```

The resulting table is too large and complex for effective understanding due to the number of numeric variables, but it highlights a high proportion of missing data (NA values).

# 3. Prediction of cGVHD outcome
## 3.1. Define DataSet

Survival Analysis are planned. 

*Event:* Diagnosis of cGVHD (cGVHD_flag == 1).

*Survival Time:*

- For patients with cGVHD: cGVHD_Diagnosis_Day as the survival time.
- For patients without cGVHD: Observation_Days_Count as the censored survival time.

*Perform PCA to reduce 167 variables into fewer components that explain the maximum variance!*

``` {r define dataset Prediction of cGVHD outcome}

desired_suffixes <- c("_30", "_60", "_90", "_180", "_365", "_cGVHD")

# include columns ending with suffixes
include_pattern <- paste0("(", paste0(desired_suffixes, collapse = "|"), ")$")

# exclude columns like _90cGVHD, _60cGVHD, etc.
exclude_pattern <- "_[0-9]+_cGVHD$"

# Select columns that match the include pattern and do not match the exclude pattern
df_cGVHD_prediction <- df_transformed %>%
  select(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, 
         matches(include_pattern)) %>%
  select(-matches(exclude_pattern))
```

**Crucial steps:**

1. The dataset (df_cGVHD_prediction) includes immune cell measurements at specific time points (_30, _60, _90, _180, _365, and _cGVHD) while excluding those associated with timing after chronic GVHD (e.g., _90_cGVHD).

```{r visualisation after choosing necessary vars}
initial_cols <- ncol(df_transformed)
final_cols <- ncol(df_cGVHD_prediction)

reduction_data <- data.frame(
  Step = c("Initial Variables", "After time points were choosen "),
  Variables = c(initial_cols, final_cols)
)

ggplot(reduction_data, aes(x = Step, y = Variables, fill = Step)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Variables), vjust = 1.5, size = 4) +
  scale_fill_manual(values = c("#1f78b4", "#33a02c")) +
  theme_minimal() +
  labs(
    title = "Reduction of Variables Due to Missing Data",
    x = "Step",
    y = "Number of Variables"
  )

```

## 3.2. Check for Missing Data

```{r check for missing data}

remove_cols_with_na <- function(data, threshold = 20, always_keep = c("Subject_ID", "Observation_Days_Count", "cGVHD_Diagnosis_Day", "cGVHD_flag")) {
  # Calculate missing values
  na_percent <- colMeans(is.na(data)) * 100

  cols_to_keep <- names(na_percent)[na_percent <= threshold]
  cols_to_keep <- unique(c(cols_to_keep, always_keep))

  cols_to_keep <- intersect(cols_to_keep, names(data))
  
  # keep only the desired columns
  cleaned_data <- data[, cols_to_keep, drop = FALSE]
  
  return(cleaned_data)
}

# Apply the function
df_cGVHD_prediction_clean <- remove_cols_with_na(
  data = df_cGVHD_prediction
)

```

**Crucial steps:**

1. Columns with more than *20%* missing values were removed to ensure data quality (from 998 to 167 variables).

```{r visualisation after missing data deleting}
initial_cols <- ncol(df_cGVHD_prediction)
final_cols <- ncol(df_cGVHD_prediction_clean)

reduction_data <- data.frame(
  Step = c("Initial Variables", "After Removing High NA"),
  Variables = c(initial_cols, final_cols)
)

ggplot(reduction_data, aes(x = Step, y = Variables, fill = Step)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Variables), vjust = 1.5, size = 4) +
  scale_fill_manual(values = c("#1f78b4", "#33a02c")) +
  theme_minimal() +
  labs(
    title = "Reduction of Variables Due to Missing Data",
    x = "Step",
    y = "Number of Variables"
  )

```

# 3.4. Visualisation

```{r visualisation}

# Select relevant numeric columns 
correlation_data <- df_cGVHD_prediction_clean %>%
  select(where(is.numeric), cGVHD_Diagnosis_Day, Observation_Days_Count) %>%
  drop_na()  # Drop rows with NA values to compute correlations

# Compute correlation matrix
cor_matrix <- cor(correlation_data, use = "complete.obs")

# Hierarchical clustering and heatmap
pheatmap(cor_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         display_numbers = FALSE,
         fontsize_row = 2,
         fontsize_col = 2,
         main = "Clustered Heatmap for Survival Variables",
         width = 20, 
        height = 20  )

```

# 4. Diagnostics of cGVHD

Comparison on the day of the onset of chronic GVHD with the closest points in terms of timing in patients without chronic GVHD.

*Group 1:* Patients diagnosed with cGVHD (cGVHD_flag = 1).
*Group 2: *Patients without cGVHD (cGVHD_flag = 0).

- For Group 1, will be used data from the day of cGVHD onset (cGVHD_Diagnosis_Day).
- For Group 2, the closest time point to the median diagnosis day of Group 1 (Median_Diagnosis_Day) is 180 days.

Due to large ammount of immune cell markers (332):

1. Cluster immune cell markers to identify groups of similar variables.
2. After clustering, perform statistical comparisons for clasters.

## 4.1. Define Data

``` {r Diagnostics of cGVHD choose the second group }
# Plot the density plot for cGVHD Diagnosis Day
df_transformed %>%
  filter(cGVHD_flag == 1) %>%
  ggplot(aes(x = cGVHD_Diagnosis_Day)) +
  geom_density(fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Distribution of cGVHD Diagnosis Day",
    x = "cGVHD Diagnosis Day",
    y = "Density"
  )

# remove outliers using the three-sigma rule
filtered_diagnosis_day <- df_transformed %>%
  filter(cGVHD_flag == 1) %>%
  select(Subject_ID, cGVHD_Diagnosis_Day)

# Calculate mean and standard deviation
mean_value <- mean(filtered_diagnosis_day$cGVHD_Diagnosis_Day, na.rm = TRUE)
sd_value <- sd(filtered_diagnosis_day$cGVHD_Diagnosis_Day, na.rm = TRUE)

lower_bound <- mean_value - 3 * sd_value
upper_bound <- mean_value + 3 * sd_value

# Remove outliers and create a new dataframe
df_diagnostic <- df_transformed %>%
  filter(
    cGVHD_flag == 1 & 
    cGVHD_Diagnosis_Day >= lower_bound & 
    cGVHD_Diagnosis_Day <= upper_bound
  )

cat("Number of removed outliers:", nrow(df_transformed %>% filter(cGVHD_flag == 1)) - nrow(df_diagnostic), "\n")

# Calculate the median of the cleaned data
median_diagnosis_day <- median(df_diagnostic$cGVHD_Diagnosis_Day, na.rm = TRUE)
cat("Median Diagnosis Day after removing outliers:", median_diagnosis_day, "\n")

# Regex pattern to include columns ending with desired suffixes
include_pattern <- paste0("(", paste0(c("_180", "_cGVHD"), collapse = "|"), ")$")

# Regex pattern to exclude columns like _90cGVHD, _60cGVHD, etc.
exclude_pattern <- "_[0-9]+_cGVHD$"

# Select columns that match the include pattern and do not match the exclude pattern
df_diagnostic <- df_diagnostic %>%
  select(Subject_ID, Observation_Days_Count, cGVHD_Diagnosis_Day, cGVHD_flag, 
         matches(include_pattern)) %>%
  select(-matches(exclude_pattern))
```

## 4.2 Check for Missing Data

``` {r Check for Missing Data diagnostic}
# Apply the function to your dataset
df_diagnostic_clean <- remove_cols_with_na(
  data = df_diagnostic
)

```

**Crucial steps:**

1. Columns with more than *20%* missing values were removed to ensure data quality (from 336 to 168 variables).

# 5. Immunological Profiles of Patients with or without cGVHD


